---
title: "envs193ds-homework-03"
author: "Valeria Fierros"
date: "2026-03-05"
format: 
  html:
    toc: true
execute: 
  message: false
  warning: false
---

[link to homework 3 GitHub repository](https://github.com/valeriafierros/ENVS-193DS_homework-03)

# 1. Set up

```{r}
#| label: set up

#reading in packages here
library(tidyverse)
library(here)
library(janitor)
library(readxl)

#reading in data here and making new object
salinity <- read_csv("data/salinity-pickleweed.csv")
steps <- read_csv("data/personaldata-updated.csv")

```


# 2. Problems

## Problem 1. Sough soil salinity

-   electrical conductivity (mS/cm)

### a.  An appropriate test

<!-- -->

1.  simple linear regression (linear model --\> lm())

-   describes a linear relationship between response (California pickleweed biomass) and predictor variables (salinity).

2.  Pearson's correlation

-   measures the strength and direction of a linear association (r)

The simple linear regression will tell us how salinity affects or predicts plant biomass, whereas the Pearson's correlation will tell us exactly *how* plant biomass and salinity are associated with one another.

### b. Creating a visualization

```{r}
#| label: salinity and biomass plot

#base layer: ggplot
ggplot(data = salinity,
       aes (x = salinity_mS_cm,
            y = pickleweed)) +
  #adding points and making them forest green
  geom_point(color = "forestgreen") +
  #adding a trendline
  geom_smooth(method = "lm",
              se = FALSE)+
  #adding labels and title
  labs(title = "As soil salinity (mS/cm) increases, California pickleweed biomass also \n increases.",
       x = "Soil Salinity (mS/cm)",
       y = "California Pickleweed Biomass (g)")+
  #changing default theme
  theme_light()


```

### c. Checking assumptions + running test (pearson correlation)

#### check your assumptions
```{r}
#| label: checking visual assumptions with scatterplot (linearity)

#base layer: ggplot
ggplot(data = salinity,
       aes (x = salinity_mS_cm,
            y = pickleweed)) +
  #adding points and making them forest green
  geom_point(color = "forestgreen") +
  #adding a trendline
  geom_smooth(method = "lm",
              se = FALSE)+
  #adding labels and title
  labs(title = "As soil salinity (mS/cm) increases, California pickleweed biomass also \n increases.",
       x = "Soil Salinity (mS/cm)",
       y = "California Pickleweed Biomass (g)")+
  #changing default theme
  theme_light()


```

```{r}
#| label: checking normality w qqplot and residuals
#| fig-height: 6
#| fig-width: 8

#making a model for qqplot 
salinity_model <- lm(pickleweed ~ salinity_mS_cm,
                     data = salinity)

#plotting salinity model to check QQplot
par(mfrow = c(2,2))
plot(salinity_model)
```

I checked the assumptions of linearity and normality of residuals (and assumed that the data is independent). The linearity assumption was evaluated using a scatterplot, and the normality of the data was evaluated using a QQplot. Based on these plots, we can conclude that the data is approximately linear (not many points in the scatterplot, but they are monotonic) and normal (no significant outliers or deviations in the qqplot). Therefore, we can say that the assumptions are reasonably met. 

#### run your test
```{r}

#running pearson's correlation --> describes the strength of the relationship between salinity and biomass with directionality
cor.test(salinity$pickleweed, salinity$salinity_mS_cm,
         method = "pearson") 

```

### d. results communication
Results: The results of our pearson's correlation test indicate that there is a moderate, positive linear relationship between salinity (mS/cm) and California pickleweed biomass. The correlation coefficient (r = 0.534) indicates a moderate positive correlation, meaning that pickleweed biomass tends to increase as salinity increases. The p-value ( p = 0.0086) is less than 0.05, so the relationship is statistically significant, and we can reject the null hypothesis that the true correlation is 0 (or that there is no correlation). The 95% CI [0.16, 0.78] does not include 0, further supporting the positive relationship between the response and predictor variables. 
